---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spatialgridr

<!-- badges: start -->
<!-- badges: end -->

`spatialgridr` provides functions for gridding spatial data; i.e. taking raw spatial data and getting that data into a grid.

## Installation

You can install the development version of spatialgridr from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("emlab-ucsb/spatialgridr")
```

`spatialgridr` has two functions:

* `get_grid()`: for creating a spatial grid
* `get_data_in_grid()`: for gridding spatial data


## Example

This shows how to obtain a spatial grid and grid some data using that grid.

```{r include=FALSE}
#install from Github if needed
remotes::install_github("emlab-ucsb/spatialgridr")
library(spatialgridr)
```

We can obtain grids in raster (`terra::rast`) or vector (`sf`) format. First we need a polygon that we want to create a grid for. In this example we will use the exclusive economic zone (EEZ) of Samoa. We also need to provide a suitable projection for the area we are interested in, https://projectionwizard.org is useful for this purpose. For spatial planning, equal area projections are normally best.

```{r grid_raster}
#load Samoa EEZ
samoa_eez <- readRDS(system.file("extdata", "samoa_eez.rds", package = "spatialgridr"))

#equal area projection for Samoa obtained from https://projectionwizard.org
samoa_projection <- '+proj=laea +lon_0=-172.5 +lat_0=0 +datum=WGS84 +units=m +no_defs'

# Create a raster grid with 10km sized cells
planning_grid <- get_grid(area_polygon = samoa_eez, projection_crs = samoa_projection, resolution = 10000)

#plot the grid
terra::plot(planning_grid)
terra::lines(terra::as.polygons(planning_grid, dissolve = FALSE)) #add the outlines of each cell

```

To obtain a grid in `sf` format we can use arguments `option = "sf_square"` or `option = "sf_hex"` in `get_grid` to specify square or hexagonal cells. We will create and plot a hexagonal grid with 10 km wide cells.

```{r grid_sf}

planning_grid_sf <- get_grid(area_polygon = samoa_eez, projection_crs = samoa_projection, resolution = 10000, option = "sf_hex")

plot(planning_grid_sf)
```

Now we can grid some data. Data can be in raster (`terra::rast()`) or `sf` format. Here's an example using global data mapping knolls (small seamounts) which is in `sf` format:

```{r grid_sf_data, warning=FALSE}
# knolls data for area of Pacific
knolls <- readRDS(system.file("extdata", "ridges.rds", package = "spatialgridr"))

#grid the data
knolls_gridded <- get_data_in_grid(spatial_grid = planning_grid, dat = knolls)

#plot
terra::plot(knolls_gridded)
terra::lines(samoa_eez |> sf::st_transform(crs = samoa_projection)) #add Samoa's EEZ
```

And another example using raster data, in this case global cold water coral distribution data which has been pre-cropped to the Pacific

```{r grid_raster_data}
#load cold water coral data
cold_coral <- terra::rast(system.file("extdata", "cold_coral.tif", package = "spatialgridr"))

#grid the data
coral_gridded <- get_data_in_grid(spatial_grid = planning_grid, dat = cold_coral)

#plot
terra::plot(coral_gridded)
terra::lines(samoa_eez |> sf::st_transform(crs = samoa_projection)) #add Samoa's EEZ
```



